\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-01-23 paper.qmd
Rewrite the DR section

To create low-dimensional representations (typically in 2D) (m-vis) [@article59] of high-D data, it is common to apply dimension reduction (DR) techniques. Approaches for DR involve linear methods such as principal component analysis (PCA) [@Karl1901], non-linear methods such as multi-dimensional scaling (MDS) [@Torgerson1967]. In the past decade, many new non-linear dimension reduction (NLDR) techniques have emerged, such as t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008] and uniform manifold approximation and projection (UMAP) [@Leland2018], designed to capture the complex and non-linear relationships present within high-D data [@Johnstone2009]. NLDR techniques are the 2D models of high-D data in our context.

It is important to visualize various Non-Linear Dimensionality Reduction (NLDR) techniques for the same high-D data in order to understand and find the best representation. After doing so, the 2D models may differ considerably from each other and may also deviate from the original data structure in high-dimensional space. Therefore, visualizing the 2D model in high-D space (m-in-ds) is more useful to answer different types of questions:

@@

Consider the high-dimensional data a rectangular matrix $X_{n \times p}$, where $X_{n \times p} = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top$, with $n$ observations in $p$ dimensions. The objective is to discover a low-dimensional projection $Y_{n \times d} = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top$, represented as an $n$ × $d$ matrix, where $d \ll p$. The reduction process seeks to remove noise from the original data set while retaining essential information.

There are two main categories of dimension reduction techniques: linear and non-linear methods. Linear techniques involve a linear transformation of the data, with one popular example being PCA. PCA performs an eigen-decomposition of the sample covariance matrix to obtain orthogonal principal components that capture the variance of the data [@Karl1901]. However, linear methods may not fully capture complex non-linear relationships present in the data.

In contrast, NLDR techniques generate the low-dimensional representation $Y$ from the high-dimensional data $X$, often using pre-processing techniques like $k$-nearest neighbors graph or kernel transformations. Multidimensional Scaling (MDS) is a class of NLDR methods that aims to construct an embedding $Y$ in a low-dimensional space, approximating the pair-wise distances in $X$ [@Torgerson1967]. Variants of MDS include non-metric scaling [@article62] and Isomap, which estimate geodesic distances to create the low-dimensional representation [@article63]. Other approaches based on diffusion processes, like diffusion maps [@article64] and the PHATE algorithm [@article03], also fall under NLDR methods.

A challenge with NLDR methods is selecting and tuning appropriate parameters. One specific technique we focus on is Pairwise Controlled Manifold Approximation (PaCMAP). Similar considerations apply to related methods like tSNE [@Laurens2008], UMAP [@Leland2018], and TrMAP [@article02].

It is important to note that methods like PCA and auto-encoders [@article65] provide a reverse mapping from the low-dimensional space back to the high-dimensional space, enabling data reconstruction. However, many non-linear methods, including tSNE, prioritize visualization and exploration over reconstruction. Their focus is on capturing complex structures that may not be easily represented in the original space, making a straightforward reverse mapping challenging.

@@

Non-linear dimension reduction techniques play a crucial role in the analysis and visualization of high-dimensional data, where the complexities of relationships among variables may not be adequately captured by linear methods. Among these techniques, tSNE stands out for its emphasis on preserving pairwise distances. By minimizing the divergence between probability distributions in both the high and low-dimensional spaces, t-SNE effectively reveals intricate structures and patterns within the data. Its application is widespread in tasks requiring the visualization of clusters and local relationships, though it does require careful consideration of the perplexity parameter for optimal results.

@@

In contrast, PaCMAP introduces supervised learning to create a low-dimensional representation while preserving pair-wise relationships. Constructing a graph based on pair-wise distances, PaCMAP optimizes an embedding using a customizable loss function. Particularly notable is PaCMAP's flexibility in incorporating class labels or additional information to guide the embedding process, offering users a means to customize its behavior and performance.

>>>>

To create low-dimensional representations (typically in 2D) (m-vis) [@article59] of high-D data, it is common to apply dimension reduction (DR) techniques. Approaches for DR involve linear methods such as principal component analysis (PCA) [@Karl1901], non-linear methods such as multi-dimensional scaling (MDS) [@Torgerson1967]. In the past decade, many new non-linear dimension reduction (NLDR) techniques have emerged, such as t-distributed stochastic neighbor embedding (tSNE) [@Laurens2008] and uniform manifold approximation and projection (UMAP) [@Leland2018]. NLDR techniques are the 2D models of high-D data in our context.

It is important to visualize various non-linear dimensionality reduction (NLDR) techniques for the same high-D data in order to understand and find the best representation. After doing so, the 2D models may differ considerably from each other and may also deviate from the original data structure in high-dimensional space. Therefore, visualizing the 2D model in high-D space (m-in-ds) is more useful to answer different types of questions:

@@

Consider the high-D data a rectangular matrix $X_{n \times p}$, where $X_{n \times p} = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top$, with $n$ observations in $p$ dimensions. The objective is to discover a low-dimensional projection $Y_{n \times d} = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top$, represented as an $n$ × $d$ matrix, where $d \ll p$. The reduction process seeks to remove noise from the original data set while retaining essential information.

There are two main categories of dimension reduction techniques: linear and non-linear methods. Linear techniques involve a linear transformation of the data, with one popular example being PCA. PCA performs an eigen-decomposition of the sample covariance matrix to obtain orthogonal principal components that capture the variance of the data [@Karl1901].

In contrast, NLDR techniques generate the low-dimensional representation $Y$ from the high-dimensional data $X$, often using pre-processing techniques like $k$-nearest neighbors graph or kernel transformations. Multidimensional Scaling (MDS) is a class of NLDR methods that aims to construct an embedding $Y$ in a low-dimensional space, approximating the pair-wise distances in $X$ [@Torgerson1967]. Variants of MDS include non-metric scaling [@article62] and Isomap, which estimate geodesic distances to create the low-dimensional representation [@article63]. Other approaches based on diffusion processes, like diffusion maps [@article64] and the PHATE (Potential of Heat-diffusion for Affinity-based Trajectory Embedding) algorithm [@article03], also fall under NLDR methods.

<!--Their focus is on capturing complex structures that may not be easily represented in the original space, making a straightforward reverse mapping challenging.

One specific technique we focus on is Pairwise Controlled Manifold Approximation (PaCMAP). Similar considerations apply to related methods like tSNE [@Laurens2008], UMAP [@Leland2018], and TrMAP [@article02].

However, linear methods may not fully capture complex non-linear relationships present in the data.-->

@@

NLDR techniques are crucial for analyzing and displaying high-dimensional data, where linear approaches may not adequately capture complexities in relationships between variables [@Johnstone2009]. One of the challenges with NLDR techniques is the selection and tuning of appropriate hyperparameters [@liao2023]. This process involves finding the suitable combination of hyperparameters that enhances the performance of the NLDR technique, considering the characteristics of the dataset and the specific goals of the analysis.

<!--This process requires a delicate balance to optimize the performance of NLDR techniques, taking into account the intricacies of the dataset and the specific goals of the analysis.-->

Additionally, another challenge is lack of reverse mapping. Techniques like PCA and auto-encoders [@article65] provide a way to map back from the low-dimensional space to the high-D space, facilitating data reconstruction. However, some NLDR methods, such as tSNE, don't have a specific way to reconstruct the original data from the low-dimensional space.

In this article, mainly focus on five NLDR techniques. They are tSNE, UMAP, PHATE, TriMAP [@article02], and Pairwise Controlled Manifold Approximation (PaCMAP) [@article01].

Among these, tSNE [@Laurens2008] stands out for its ability to preserve pairwise distances. By minimizing the divergence between probability distributions in both high and low-dimensional spaces, tSNE effectively uncovers intricate structures and patterns within the data. Its application is widespread, particularly in tasks requiring the visualization of clusters and local relationships. However, achieving effective results requires careful consideration of hyperparameters, such as perplexity.

UMAP [@Leland2018] is a useful technique for simplifying data while maintaining both local and overall structures. It builds a fuzzy topological view by considering nearby data points and then optimizes a simplified version to match that view. UMAP is known for working well with different scales of relationships in data and is efficient in handling large datasets. However, it's important to choose parameters like neighbors and minimum distance carefully, as they can affect the results.

Furthermore, PHATE [@article03] is great for understanding how things develop, especially in single-cell genomics. It uses a heat diffusion process to capture relationships between data points, like points along a trajectory. While PHATE is excellent for revealing these developmental structures, it requires careful tuning of its parameters because of its specialized focus.

Additionally, TriMAP [@article02] takes a special approach by creating a triangulated graph representation of the data. This method is good at understanding both local and global structures by treating the data as a network of triangles. TriMAP is powerful in capturing complicated structures, but it's important to choose parameters carefully, like deciding how many neighbors to consider.

PaCMAP [@article01] is different because it adds supervised learning to make a 2D representation while keeping the relationships between pairs of points. It builds a graph using distances between pairs and then makes the 2D representation better using a customizable loss function. What's special about PaCMAP is that it can use class labels or extra information to guide how it makes the 2D representation. This gives users a way to change how PaCMAP works to fit their needs better.

<!--Among these techniques, tSNE stands out for its emphasis on preserving pairwise distances. By minimizing the divergence between probability distributions in both the high and low-dimensional spaces, t-SNE effectively reveals intricate structures and patterns within the data. Its application is widespread in tasks requiring the visualization of clusters and local relationships, though it does require careful consideration of the perplexity parameter for optimal results.

@@

In contrast, PaCMAP introduces supervised learning to create a low-dimensional representation while preserving pair-wise relationships. Constructing a graph based on pair-wise distances, PaCMAP optimizes an embedding using a customizable loss function. Particularly notable is PaCMAP's flexibility in incorporating class labels or additional information to guide the embedding process, offering users a means to customize its behavior and performance.-->

### 2024-01-23 paper.tex
Rewrite the DR section

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, boxrule=0pt, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

To create low-dimensional representations (typically in 2D) (m-vis)
\citep{article59} of high-D data, it is common to apply dimension
reduction (DR) techniques. Approaches for DR involve linear methods such
as principal component analysis (PCA) \citep{Karl1901}, non-linear
methods such as multi-dimensional scaling (MDS) \citep{Torgerson1967}.
In the past decade, many new non-linear dimension reduction (NLDR)
techniques have emerged, such as t-distributed stochastic neighbor
embedding (tSNE) \citep{Laurens2008} and uniform manifold approximation
and projection (UMAP) \citep{Leland2018}, designed to capture the
complex and non-linear relationships present within high-D data
\citep{Johnstone2009}. NLDR techniques are the 2D models of high-D data
in our context.

It is important to visualize various Non-Linear Dimensionality Reduction
(NLDR) techniques for the same high-D data in order to understand and
find the best representation. After doing so, the 2D models may differ
considerably from each other and may also deviate from the original data
structure in high-dimensional space. Therefore, visualizing the 2D model
in high-D space (m-in-ds) is more useful to answer different types of
questions:

@@

Consider the high-dimensional data a rectangular matrix
\(X_{n \times p}\), where
\(X_{n \times p} = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top\),
with \(n\) observations in \(p\) dimensions. The objective is to
discover a low-dimensional projection
\(Y_{n \times d} = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top\),
represented as an \(n\) × \(d\) matrix, where \(d \ll p\). The reduction
process seeks to remove noise from the original data set while retaining
essential information.

There are two main categories of dimension reduction techniques: linear
and non-linear methods. Linear techniques involve a linear
transformation of the data, with one popular example being PCA. PCA
performs an eigen-decomposition of the sample covariance matrix to
obtain orthogonal principal components that capture the variance of the
data \citep{Karl1901}. However, linear methods may not fully capture
complex non-linear relationships present in the data.

In contrast, NLDR techniques generate the low-dimensional representation
\(Y\) from the high-dimensional data \(X\), often using pre-processing
techniques like \(k\)-nearest neighbors graph or kernel transformations.
Multidimensional Scaling (MDS) is a class of NLDR methods that aims to
construct an embedding \(Y\) in a low-dimensional space, approximating
the pair-wise distances in \(X\) \citep{Torgerson1967}. Variants of MDS
include non-metric scaling \citep{article62} and Isomap, which estimate
geodesic distances to create the low-dimensional representation
\citep{article63}. Other approaches based on diffusion processes, like
diffusion maps \citep{article64} and the PHATE algorithm
\citep{article03}, also fall under NLDR methods.

A challenge with NLDR methods is selecting and tuning appropriate
parameters. One specific technique we focus on is Pairwise Controlled
Manifold Approximation (PaCMAP). Similar considerations apply to related
methods like tSNE \citep{Laurens2008}, UMAP \citep{Leland2018}, and
TrMAP \citep{article02}.

It is important to note that methods like PCA and auto-encoders
\citep{article65} provide a reverse mapping from the low-dimensional
space back to the high-dimensional space, enabling data reconstruction.
However, many non-linear methods, including tSNE, prioritize
visualization and exploration over reconstruction. Their focus is on
capturing complex structures that may not be easily represented in the
original space, making a straightforward reverse mapping challenging.

@@

Non-linear dimension reduction techniques play a crucial role in the
analysis and visualization of high-dimensional data, where the
complexities of relationships among variables may not be adequately
captured by linear methods. Among these techniques, tSNE stands out for
its emphasis on preserving pairwise distances. By minimizing the
divergence between probability distributions in both the high and
low-dimensional spaces, t-SNE effectively reveals intricate structures
and patterns within the data. Its application is widespread in tasks
requiring the visualization of clusters and local relationships, though
it does require careful consideration of the perplexity parameter for
optimal results.

UMAP is another powerful non-linear technique that strikes a balance
between preserving local and global structures. Constructing a fuzzy
topological representation using a weighted k-nearest neighbors graph,
UMAP optimizes the low-dimensional embedding to resemble this
representation. Known for its efficiency and scalability, UMAP is
versatile across various scales of relationships in the data, although
parameter sensitivity, particularly concerning the choice of neighbors,
must be taken into account.

For trajectory data, PHATE provides specialized capabilities. It models
the affinity between data points, simulating a heat diffusion process to
capture developmental processes, particularly in single-cell genomics.
While PHATE excels in revealing trajectory structures and offering
insights into cellular development, it necessitates careful parameter
tuning due to its specialized nature.

TriMAP adopts a unique approach by approximating the data manifold
through the construction of a triangulated graph representation. This
technique efficiently captures both global and local structures by
representing the data as a network of triangles. TriMAP's strength lies
in its ability to efficiently capture complex structures, albeit with
sensitivity to parameter choices, including the number of neighbors.

In contrast, PaCMAP introduces supervised learning to create a
low-dimensional representation while preserving pair-wise relationships.
Constructing a graph based on pair-wise distances, PaCMAP optimizes an
embedding using a customizable loss function. Particularly notable is
PaCMAP's flexibility in incorporating class labels or additional
information to guide the embedding process, offering users a means to
customize its behavior and performance.

>>>>

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, breakable, interior hidden, enhanced, boxrule=0pt, frame hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

To create low-dimensional representations (typically in 2D) (m-vis)
\citep{article59} of high-D data, it is common to apply dimension
reduction (DR) techniques. Approaches for DR involve linear methods such
as principal component analysis (PCA) \citep{Karl1901}, non-linear
methods such as multi-dimensional scaling (MDS) \citep{Torgerson1967}.
In the past decade, many new non-linear dimension reduction (NLDR)
techniques have emerged, such as t-distributed stochastic neighbor
embedding (tSNE) \citep{Laurens2008} and uniform manifold approximation
and projection (UMAP) \citep{Leland2018}. NLDR techniques are the 2D
models of high-D data in our context.

It is important to visualize various non-linear dimensionality reduction
(NLDR) techniques for the same high-D data in order to understand and
find the best representation. After doing so, the 2D models may differ
considerably from each other and may also deviate from the original data
structure in high-dimensional space. Therefore, visualizing the 2D model
in high-D space (m-in-ds) is more useful to answer different types of
questions:

@@

Consider the high-D data a rectangular matrix \(X_{n \times p}\), where
\(X_{n \times p} = \begin{bmatrix} \textbf{x}_{1} & \textbf{x}_{2} & \cdots & \textbf{x}_{n}\\ \end{bmatrix}^\top\),
with \(n\) observations in \(p\) dimensions. The objective is to
discover a low-dimensional projection
\(Y_{n \times d} = \begin{bmatrix} \textbf{y}_{1} & \textbf{y}_{2} & \cdots & \textbf{y}_{n}\\ \end{bmatrix}^\top\),
represented as an \(n\) × \(d\) matrix, where \(d \ll p\). The reduction
process seeks to remove noise from the original data set while retaining
essential information.

There are two main categories of dimension reduction techniques: linear
and non-linear methods. Linear techniques involve a linear
transformation of the data, with one popular example being PCA. PCA
performs an eigen-decomposition of the sample covariance matrix to
obtain orthogonal principal components that capture the variance of the
data \citep{Karl1901}.

In contrast, NLDR techniques generate the low-dimensional representation
\(Y\) from the high-dimensional data \(X\), often using pre-processing
techniques like \(k\)-nearest neighbors graph or kernel transformations.
Multidimensional Scaling (MDS) is a class of NLDR methods that aims to
construct an embedding \(Y\) in a low-dimensional space, approximating
the pair-wise distances in \(X\) \citep{Torgerson1967}. Variants of MDS
include non-metric scaling \citep{article62} and Isomap, which estimate
geodesic distances to create the low-dimensional representation
\citep{article63}. Other approaches based on diffusion processes, like
diffusion maps \citep{article64} and the PHATE (Potential of
Heat-diffusion for Affinity-based Trajectory Embedding) algorithm
\citep{article03}, also fall under NLDR methods.

@@

NLDR techniques are crucial for analyzing and displaying
high-dimensional data, where linear approaches may not adequately
capture complexities in relationships between variables
\citep{Johnstone2009}. One of the challenges with NLDR techniques is the
selection and tuning of appropriate hyperparameters \citep{liao2023}.
This process involves finding the suitable combination of
hyperparameters that enhances the performance of the NLDR technique,
considering the characteristics of the dataset and the specific goals of
the analysis.

Additionally, another challenge is lack of reverse mapping. Techniques
like PCA and auto-encoders \citep{article65} provide a way to map back
from the low-dimensional space to the high-D space, facilitating data
reconstruction. However, some NLDR methods, such as tSNE, don't have a
specific way to reconstruct the original data from the low-dimensional
space.

In this article, mainly focus on five NLDR techniques. They are tSNE,
UMAP, PHATE, TriMAP \citep{article02}, and Pairwise Controlled Manifold
Approximation (PaCMAP) \citep{article01}.

Among these, tSNE \citep{Laurens2008} stands out for its ability to
preserve pairwise distances. By minimizing the divergence between
probability distributions in both high and low-dimensional spaces, tSNE
effectively uncovers intricate structures and patterns within the data.
Its application is widespread, particularly in tasks requiring the
visualization of clusters and local relationships. However, achieving
effective results requires careful consideration of hyperparameters,
such as perplexity.

UMAP \citep{Leland2018} is a useful technique for simplifying data while
maintaining both local and overall structures. It builds a fuzzy
topological view by considering nearby data points and then optimizes a
simplified version to match that view. UMAP is known for working well
with different scales of relationships in data and is efficient in
handling large datasets. However, it's important to choose parameters
like neighbors and minimum distance carefully, as they can affect the
results.

Furthermore, PHATE \citep{article03} is great for understanding how
things develop, especially in single-cell genomics. It uses a heat
diffusion process to capture relationships between data points, like
points along a trajectory. While PHATE is excellent for revealing these
developmental structures, it requires careful tuning of its parameters
because of its specialized focus.

Additionally, TriMAP \citep{article02} takes a special approach by
creating a triangulated graph representation of the data. This method is
good at understanding both local and global structures by treating the
data as a network of triangles. TriMAP is powerful in capturing
complicated structures, but it's important to choose parameters
carefully, like deciding how many neighbors to consider.

PaCMAP \citep{article01} is different because it adds supervised
learning to make a 2D representation while keeping the relationships
between pairs of points. It builds a graph using distances between pairs
and then makes the 2D representation better using a customizable loss
function. What's special about PaCMAP is that it can use class labels or
extra information to guide how it makes the 2D representation. This
gives users a way to change how PaCMAP works to fit their needs better.
